Modelling BCG vaccination in the UK

### 2019-08-05 rmds/04-tb-epi-england.Rmd
updated measurements param table

For date of symptom onset there was strong evidence of an MNAR mechanism for all variables considered, except for sex (Table \@ref(tab:symptonset-miss)). As found previously, the likelihood of date of symptom onset being missing reduced with year of notification. Children (0-14 years old) were more likely to have a missing date of symptom onset than any other age group as were those in any socio-economic quintile when compared to the poorest group. UK born cases were more likely to have a complete date of symptom onset than non-UK born cases, with the White ethnic group being more likely to have a missing date of symptom onset than most other ethnic groups.

>>>>

For date of symptom onset there was strong evidence of an MAR mechanism for all variables considered, except for sex (Table \@ref(tab:symptonset-miss)). As found previously, the likelihood of date of symptom onset being missing reduced with year of notification. Children (0-14 years old) were more likely to have a missing date of symptom onset than any other age group as were those in any socio-economic quintile when compared to the poorest group. UK born cases were more likely to have a complete date of symptom onset than non-UK born cases, with the White ethnic group being more likely to have a missing date of symptom onset than most other ethnic groups.

### 2019-08-05 rmds/09-model-fitting.Rmd
updated measurements param table

First, I will detail the data that will be used to fit the state-space model, then go on to describe the observation model that maps the model results to the observed data.

The primary data source for the model is the reported, age-stratified, UK born TB notifications from 2000 to 2004 as recorded in the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). 2000 to 2004 are the years for which notifications are stratified by UK birth status with universal school-age BCG vaccination. Data was grouped using the age groups present in the dynamic model (5 year age groups up to 49 and then a group from 50-69 and a 70+ group). Using the age-stratified incidence data allows for more complex trends to be identified.

Additional datasets were considerd during initial model fitting and duringl later model calibartion. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. Reducing the impact of stochastic noise, which makes fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were used as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. The downside of using this data is that twofold. Firstly, reporting standards may have changed over time so a single measurement model may not be appropriate. Secondly, non-UK born cases are included in these data making fitting to this data very dependent on

>>>>

The primary data source for the model is the reported, age-stratified, UK born TB notifications from 2000 to 2004 as recorded in the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). 2000 to 2004 are the years for which notifications are stratified by UK birth status with universal school-age BCG vaccination. Data were grouped using the age groups present in the dynamic model (5 year age groups up to 49 and then a group from 50-69 and a 70+ group). Using age-stratified incidence data, versus aggregated data, allows for more complex trends to be identified.

Additional datasets were considered during initial model fitting and during later model calibration. These were a condensed version of the age stratified data discussed above with a reduced number of age groups (children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old)) and a dataset of historic pulmonary TB notifications. The advantage of condensing age groups was that the number of notifications in each group increased. Reducing the impact of stochastic noise, which makes fitting the model easier as trends in the data are more consistent. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approach used was highly compute intensive. The downside of this approach is that potentially important information may be lost when data is condensed into fewer groups. Historic pulmonary TB notifications (including both UK born and non-UK born cases) from 1990, 1994, and 1998 were considered as using data from the decade prior to the time period of interest allows the long term trends to be fitted to. A subset of the available data was used as this limited the impact on the compute time of the fitting pipeline. These data were originally collected in the Statutory Notifications of Infectious Diseases (NOIDS) dataset with notifications from 1913 to 1999. These data were sourced from Public Health England,[@PHE2017] and made available in `R` using `tbinenglanddataclean`^[Historic TB notification data via `tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/]. Downsides of using these data are that reporting standards may have changed over time so a single measurement model may not be appropriate and non-UK born cases are included in these data making fitting to this data dependent on the number of non-UK born cases pre 2000 which are themselves estimated during model fitting.
